Inference using context-dependent models of mutation

### 2021-10-06 writeups/context-dependence.tex
dropping addressed comment

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
%EM The full exact likelihood? Isn't it an approximation?
and that something will be sufficient to compute the full likelihood.
To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
%EM Can we say initial subsequence or something?
the probability that an initial sequence $x$ of length $n$, after time $t$,
is found to match a smaller sequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
%EM Say $a$ is the first $\ell$ bases of $x$ and $b$ is the last $r$ bases of $x$.
We refer to such patterns as ``T-mers'', as depicted in Figure~\ref{fig:Tmers}.
%EM Later we use the terminology "base of the T-mer." Let's define that here.
%EM Shall we define w here?
This quantity is found by marginalizing $p_m$:
\begin{align}
%EM Are we parameterizing in terms of n,\ell,r or n,\ell,m? We see both.
    p_{n,\ell,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
%EM should this be p_{n,\ell,m} below?
%EM What is k here? And do you mean \ell?
is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
\begin{align} \label{eqn:window_approx}
%EM I think I know what you're doing here but the $i$ seems rather lonely here. Do we think of it as something for marginalization or summation or something?
%EM Also, x on the left hand side appears to be the full sequence, while on the right hand side it appears to be a subsequence. Same for y.
    p_{n,\ell,r}(x,y) \approx \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}.
\end{align}

>>>>

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
%EM The full exact likelihood? Isn't it an approximation?
and that something will be sufficient to compute the full likelihood.
To do this, let $p_{n,\ell,r}(x,y)$ for $r+\ell < n$ denote
the probability that an initial sequence $x$ of length $n$, after time $t$,
is found to match a smaller sequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
%EM Say $a$ is the first $\ell$ bases of $x$ and $b$ is the last $r$ bases of $x$.
We refer to such patterns as ``T-mers'', as depicted in Figure~\ref{fig:Tmers}.
%EM Later we use the terminology "base of the T-mer." Let's define that here.
%EM Shall we define w here?
This quantity is found by marginalizing $p_m$:
\begin{align}
%EM Are we parameterizing in terms of n,\ell,r or n,\ell,m? We see both.
    p_{n,\ell,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_n(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^n, \; y \in \calS^{n-\ell-r} ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
We expect $p_{n,\ell,m}(x,y)$ to be asymptotically correct when $\ell$ is large (and $t$ is fixed),
i.e.\ that the probability a given subsequence matching $x$ is later observed to have its center matching $y$
%EM should this be p_{n,\ell,m} below?
%EM What is k here? And do you mean \ell?
is close to $p_{\ell,m}(x,y)$, regardless of context wider than $2k+l$:
\begin{align} \label{eqn:window_approx}
%EM I think I know what you're doing here but the $i$ seems rather lonely here. Do we think of it as something for marginalization or summation or something?
%EM Also, x on the left hand side appears to be the full sequence, while on the right hand side it appears to be a subsequence. Same for y.
    p_{n,\ell,r}(x,y) \approx \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}.
\end{align}
